---
title: "Module 01 Lab 01"
author: "HD Sheets"
date: "2024-10-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, results='hide')
options(max.print = 20) 
```

## Module 01 Lab 01

For each problem below,  create the cells needed and enter your solution

# Problem 1

a.) Use seq to create a vector of values x from 10 to 1, decreasing in steps of 1/2
```{r}
x=seq(from=10, to=1, by=(-0.5))
x
```

b.) Find the length of this vector, using an R function
```{r}
length(x)
```

c.) Find the index of the element in x equal to 7.5
```{r}
index_1<-which(x==7.5)
index_1
```

d.) Remove the value 7.5 from the vector x,  so the length decreases by 1
```{r}
x<-x[-index_1]
x
```

e.) Sort the vector into both increasing form and decreasing form
```{r}
sort(x,decreasing=FALSE)
sort(x,decreasing=TRUE)
```

# Problem 2

Read the help manual for the function runif()

a.) Generate a vector z of 10 random uniform values between 1 and 5
```{r}
z<-runif(10,1,5)
z
```

b.) Find the minimum and maximum
```{r}
min(z)
max(z)
```

c.) Find the indices of the minimum and maximum
```{r}
min_index<-which(z==min(z))
min_index
max_index<-which(z==max(z))
max_index
```

d.) Compute the vector of the square values in z, store this in a vector called
     z2
```{r}
z2<-(z^2)
z2
```

e.) Find out if any of the squared vectors are less than 4
```{r}
any(z2<4)
```


f.) Find out of all the squared vectors are less than 22
```{r}
all(z2<22)
```

g.) find out how many  of the squared vectors are greater than 15
      (hint,  try summing the test statement)
```{r}
sum(z2>15)
```
    

# Problem 3

Read the help manual for rnorm to generate random normal values

a.) create a vector of 9 values, all of which are equal to 3
```{r}
nine_values<-rnorm(9, mean = 3, sd = 0)
nine_values
```

b.) Put this into a 3 by 3 matrix
```{r}
matrix=matrix(nine_values,3,3)
matrix
```

c.) Label the columns as "a", "b","c", and label the rows 1 to 3
```{r}
colnames(matrix)=c("a","b","c")
rownames(matrix)=c("1","2","3")
```
d.) create a vector of 12 random normal values
```{r}
twelve_values<-rnorm(12,mean=0,sd=1)
twelve_values
```
e.) put the random normal values into  4 x 3 matrix
```{r}
matrix2=matrix(twelve_values,4,3)
matrix2

```
f.) Create a 3 x 3 identity matrix,  you will need to look up how to do this 
     in R,   google search "R identity matrix"
```{r}
ident_matrix<-diag(3)
ident_matrix
```
g.) Multiply the matrix from part b by the identify matrix created in part f
```{r}
ident_matrix%*%matrix
matrix%*%ident_matrix
```
h.) find the row and column numbers of the largest value in the matrix created in 
     step e
```{r}
dim(matrix2)
max_matrix2<-max(matrix2)
max_indices<-which(matrix2 == max_matrix2, arr.ind = TRUE)
max_indices
```
i.) What is the average value of the 2nd column of the matrix from step e
```{r}
matrix2[,2]
mean(matrix2[,2])
```
j.) Multiply the matrix from part e by the matrix from part b
```{r}
matrix2%*%matrix
```
# Problem 4

Explain the difference between vectors and lists in R

-how are each created?
# vectors are typically created with the "c()" or "combine" function; this allows you to create a set of like values c(1,2,3,4,5). In swirl practice problems, they showed creation of vectors using seq and ":" as well. Lists are created with the "list()" function; this allows you to create a set of values of different data structes list1<- list(name = "katarina", favorite_numbers =c(3,33,77). 
-What is different about the two
#vectors are sets of values that are of the same data type while lists can hold multiple different objects and data structures. 
-Create a list of the values 1,2,3.  Can you muliple the list by 2?
```{r}
list1<-list(1,2,3)
list1
#list1*2
#you cannot multiply the list by 2
```
# Problem 5

a.) Create a list containing the items   "bob",  "sally", c(1,2,3,4), seq(1,5,by=2), c("jah","nah","hah") in it
```{r}
new_list<-list("bob","sally",c(1,2,3,4), seq(1,5, by=2), c("jah","nah","hah"))
new_list
```

b.) How would you reference the word "nah"?
```{r}
new_list[[5]][[2]]
```

c.) How would you refernce the y in "sally"
```{r}
y <-substr(new_list[[2]],5,5)
y
#I tried multiple times to do: y <- new_list[[2]][5] and I kept receiving NA as the output. I had to do some research to find str_sub which provides you with the opportunity to extract or replace substrings in a character vector. 
```
d.) Extract the vector 1,2,3,4 and store it in a new variable.  Find the mean
```{r}
new_vector<-new_list[[3]]
new_vector
mean(new_vector)
```

e.) Create a new list from the one in part 1 by removing "bob" and "sally" from the list
```{r}
new_list2<-new_list[-c(1,2)]
new_list2
```
# Problem 6

Use ?swiss to look at the built-in data set swiss

a.) What type of data type is this?
```{r}
?swiss
class(swiss)
#swiss is a data frame. 
```
b.) How many rows and columns does it have (use a function to determine this)
```{r}
dim(swiss)
#47 rows, 6 columns
```
c.) Use the summary() and str() functions on this object,  explain what they mean
```{r}
summary(swiss)
str(swiss)
#summary() gives an overview of what the data contains. For this data frame it shows the min, 1st Q, Median, Mean, 3rd Q, and Max of the 6 main variables. str() shows the structure of the data frame, offering the details of what each category contains. For this example, it shows the data types of the 6 categories along with its contents. 
```
e.) Print out the column of Agriculture values, but no other columns
```{r}
agr<-swiss[,"Agriculture"]
agr
```
d.) Which region has the highest education value?
```{r}
data(swiss)
edu<-swiss$Education
edu
max_edu<-max(edu)
max_index_edu<-which(swiss$Education == max_edu)
max_index_edu
region<-rownames(swiss)[max_index_edu]
region
#V. De Geneve has the highest education value. 
```
e.) Which region has the lowest infant mortality?
```{r}
data(swiss)
inf_mort<-swiss$Infant.Mortality
inf_mort
min_inf_mort<-min(inf_mort)
min_index_infm<-which(swiss$Infant.Mortality==min_inf_mort)
min_index_infm
region<-rownames(swiss)[min_index_infm]
region
#La Vallee has the lowest infant mortality. 
```
# Problem 7

a.) Use file.choose() to get the full name and file path of the example file
sales.csv that is included in the Module 02 content

-Do not use file.choose() in the RMD file,  run it in the console.  Including
file.choose() in an RMD will cause a knit errror

b.)  set infile to the path you found and load the sales.csv file

```{R}
# this is my path name for the download,  change it to the file name you get
# using file.choose()

infile="/Users/katarinadouglas-blake/Desktop/DSE5002/Module 02/sales.csv"

sales_df=read.csv(infile)
```

c.) Use head(), str() and summary() on this data set.   How many orders are there?
```{r}
head(sales_df)
str(sales_df)
summary(sales_df)
# there are 9994 orders 
```

d.) Use the unique() function on the Segment column to find out the number of 
distinct segments there are in the data
```{r}
dist_seg<-unique(sales_df$Segment)
dist_seg
#there are three distinct segments in the data. 
```

e.) How many different customers are there?
```{r}
diff_cust<-unique(sales_df$Customer.Name)
length(diff_cust)
# there are 793 different customers. 
```
f.) Find the highest profit and the index of the highest profit
```{r}
prof<-sales_df$Profit
prof[1:10]
max_prof<-max(prof)
max_prof
max_index_prof<-which(sales_df$Profit == max_prof)
max_index_prof
#the max profit is $8,399.976 at an index of [6827]
```
g.) What was the name of the customer with the highest profit order?
```{r}
prof_cust<-(sales_df$Customer.Name[max_prof])
prof_cust
#the name of the customer with the highest profit order was Sharelle Roach. 
```
h.) Set the Segment variable to be a factor
```{r}
sales_df$Segment<-factor(sales_df$Segment)
sales_df$Segment[1:10]
```
i.) use the lubridate package to set the Order.Data and Ship.Date varialbe to be
lubridate style date variables
```{r}
library(lubridate)
sales_df$Order.Date = trimws(as.character(sales_df$Order.Date))
sales_df$Ship.Date = trimws(as.character(sales_df$Ship.Date))
sales_df$Order.Date.l=mdy(sales_df$Order.Date)
sales_df$Ship.Date.l=mdy(sales_df$Ship.Date)
```
```{r}
head(sales_df)
```
j.) compute the shipping delay from each order, using order date and ship data
     what were the longest and shortest shipping delays?
```{r}
sales_df$shippingdelay<-as.numeric(sales_df$Ship.Date.l - sales_df$Order.Date.l)
longest_delay<-max(sales_df$shippingdelay, na.rm = TRUE)
                   cat("The longest shipping delay was", longest_delay, "days \n")
shortest_delay<-min(sales_df$shippingdelay, na.rm = TRUE)
                    cat("The shortest shipping delay was", shortest_delay, "days \n")
longest_delay
shortest_delay
#The longest shipping delay was 7 days.The shortest shipping delay was 0 days. 
```

k.) Were orders received on Saturday and Sunday?   Were any shipped on Sunday?
```{r}
wknd_order<-any(wday(sales_df$Order.Date.l) %in% c(6&7))
wknd_order
sunday_shipment<-any(wday(sales_df$Ship.Date.l)==7)
sunday_shipment
```


l.)  Convert the Customer.Name to all lower case

```{r}
sales_df$Customer.Name<-tolower(sales_df$Customer.Name)
sales_df$Customer.Name
```

m.) Search the customer names and determine how many contain the name John
```{r}
library(stringr)
sum(str_detect(sales_df$Customer.Name, regex("John",ignore_case = TRUE)))
# 138 customer names contain the name John. 
```


n.) How many times is the word "table" in the product names -careful,   str_detect() might find "table" in "portable",   also be careful with case
```{r}
count_Table<-sum(str_count(sales_df$Product.Name, "(?i)\\btable\\b"))
count_Table
# the word "table" is in the product names 230 times. 
```

         
o.) Use tapply to find the mean profit value, grouped by State
```{r}
mean_profit<- tapply(sales_df$Profit, sales_df$State, mean)
mean_profit
```
   
      https://www.statology.org/tapply-r/
   
p.) use tapply to find the median profit value, grouped by Segment
```{r}
median_prof <- tapply(sales_df$Profit, sales_df$Segment, median)
median_prof
```

q.) use table to get the number of orders per segment
```{r}
seg_orders<-table(sales_df$Segment)
seg_orders
#Consumers had 5191 orders, Corporate segment had 3020 order, and Home Office had 1783 orders. 
      
```

