---
title: "Dates_and_Lubridate"
author: "HD Sheets"
date: "2024-07-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Dates and the Lubridate package

Dates are sort of a constant issue in data

In the US,  it is common to write dates in forms such as

07/09/24

or

July 9, 2024

In Europe,   9/7/24 or 9 July, 2024

the date at the top of this document is written year-month-day

Humans easily interpret dates written in different formats, but computers don't at the moment (although LLMs can probably deal with this).

It is common to see dates stored as text,  so R loads them as 
character strings.

We will often want to convert these to a specialized date from so we can

-compute differences between dates
-extract the month, or the year
-determine the day of the week of a given day 
      (why would we care about this?)
-locate holidays
-determine days of the week

The package lubridate (Hadley Wickham and Co, again) does all kinds
of handy date processing

See also

https://lubridate.tidyverse.org/

lubridate also handles time data

```{R}
require("lubridate")

```
# Converting date information from text

Lubridate has a whole set of converters, that have the form

ymd(), mdy(),dmy()... depending on the order of month,day and year

```{R}
a=ymd("2024/01/12")
print(a)
class(a)

```
Note that ymd() extracted the date information and stored it in a 
variable of type "Date"

"Date" is a Class defined in lubridate, that both stores dates and has a bunch of "member functions" or "methods" that allow us to work with dates

Here's another one

```{R}
b= mdy("July 4 2024")
print(b)

```
We can now compute a difference or elapsed time between two dates, just using a subtraction

```{R}
b-a
```
```{R}
c=b-a
class(c)
3*c
```
Notice that c is a different class than a and b

we can convert a difftime to numeric

```{R}
d=as.numeric(c)
d
class(d)

```

# *Question/Action*

in the cell below create a variable "today" that has today's date
as a datetime variable

Find the number of days since the start of the year 01/01/2024,  you'll want to convert that date to an R datetime variable as well

```{R}
# your code here

```

# Obtaining the current date and time from lubridate

```{R}
x=now()
x
class(x)
```
Note that the time zone is included

We can extract various information from this date

```{R}

# basics
month(x)
year(x)
day(x)
```




```{r}
#time information
hour(x)
minute(x)
second(x)
```

```{R}
# days of the week

wday(x,label=TRUE)

```
# Changing a time zone

Simple, if we know the name of the zone

```{R}
with_tz(x,"America/Chicago")
with_tz(x,"Etc/Greenwich")
```
# OlsonNames()

will show us the list of valid time zone names

```{R}
OlsonNames()
```

# Using Lubridate on a data set

Typically when we load a data set into a dataframe,  it will have dates in string formats.

We will want to 

a.) load the data file
b.) Use head to see how the dates are formatted
c.) use the appropiate converter to convert the column to a datetime

This set of steps is going to be pretty common, let's look at it

```{R}
#use file.choose() to get the full filepath and paste the 
# file path in here

infile="G:\\My Drive\\DSE5002\\Module_2\\sales.csv"

sales_df=read.csv(infile)

```

```{R}
head(sales_df)
```
Let's convert Order.Date from a character string to
a date-time

I'm going to convert the column and then store it in sales_df$Order.Date,  replacing the original column

We could also just create a new column

Looking at the head, it is in mdy format

```{R}
# convert
sales_df$Order.Date=mdy(sales_df$Order.Date)

#use head to make sure the conversion worked
head(sales_df)

```

# Do we have any Orders on Sundays?

we can get a list of the days of the week using wday, and then use
unique to get list of distinct days

```{R}
unique(wday(sales_df$Order.Date,label=TRUE))
```

# *Question/Action*

List below three different questions you could ask about the dates in this data file?

Try to think of three questions that would have drive "business value" for your organization that are related to the date information in this file

